# Counting Process

In this section, we will cover 1) survival data structure (i.e., counting process) and 2) modeling survival data.

## Set packages and library

```{r, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(eha)
library(survival)
library(data.table)
library(flextable)
library(survminer)
library(ggfortify)
library(ggplot2)
library(data.table)
```


```{r, include=FALSE}
oldmort01 <- oldmort
```

## Example

> "The **oldmort** dataset in **eha** package contains life histories of people followed from their 60th birthday to their 100th, or until death, born between June 28, 1765 and December 31, 1820 in Skellefteå. The variable *enter* is age at start of the given interval, *exit* contains the age at the end of the interval. We need to calculate follow-up time since age 60 - 60 is subtracted from *enter* and *exit*. The variable *event* is an indicator of death at the duration given by exit." <https://www.rdocumentation.org/packages/eha/versions/2.8.5/topics/oldmort> [Göran Broström, <http://ehar.se/r/ehar2/parametric.html>]

Here are the summary of the **oldmort** dataset.

```{r}
library(eha)

oldmort01 <- oldmort

summary(oldmort01) # descriptive statistics
str(oldmort01) # structure
head(oldmort01) # preview
```

To check how this dataset is constructed, we will need to identify any duplicated *id*.

```{r}
dup01 <- data.frame(table(oldmort01$id))
dup02 <- dup01[order(-dup01$Freq), ]
```

In the following example, please check

-   When is a new record for the same *id* created?
-   What are the time-invariant variables?
-   What are the time-variant variables?
-   What does it mean by "TRUE" or "FALSE" in *event*?
-   How does the time of *enter* and *exit* connected with each other? What would happen if there is a gap between two records?

```{r}
dup03 <- oldmort01[oldmort01$id %in% c("789000771", "796001158"),  ]
dup03
```

## Practice: AddHealth Public datasets

There are many ways to construct long-form datasets with counting process. The following procedure is just one way to achieve the goal.

Here are a couple of things to construct a long-form dataset with counting process.

-   In practice, measuring outcomes, exposures, confounders, and other variables involves a separate procedure for each one of variables. I personally prefer to divide each measurement as time-variant and time-invariant datasets, respectively.
-   Two variables should be **ALWAYS** included in every single dataset you are working on - *AID* and *wave* (or any other *Time* variable).
-   Datasets with time-invariant variables can be merged by *AID*, while those with time-variant variables need to be merged by *AID* and *wave*.
-   Time-varying variables will be assigned a single variable name. Let's say we are to use self-rated health with a variable name of *SRH* for five waves. The dataset should contain *AID*, *wave*, and *SRH*. The SRH in each wave should be assigned the same name, *SRH*, and the wave information will be on *wave*. This way, you can simply "stack up" all 5-wave data to construct the long-form datasets.

First, each *rda* dataset will be loaded and then saved as **WAVE0X**. After assigning a *wave* variable for each of them, we will keep the **WAVE0X** dataset and *WX* datasets only.


```{r}
#1st wave
load("~/NAS_Share/HP/AddHealthPublic/ICPSR_21600/DS0001/21600-0001-Data.rda")
wave01 <- da21600.0001
wave01$wave <- 1
rm(da21600.0001)
w1 = subset(wave01, select = c(AID, wave))

#2nd wave
load("~/NAS_Share/HP/AddHealthPublic/ICPSR_21600/DS0005/21600-0005-Data.rda")
wave02 <- da21600.0005
wave02$wave <- 2
rm(da21600.0005)
w2 = subset(wave02, select = c(AID, wave))

#3rd wave
load("~/NAS_Share/HP/AddHealthPublic/ICPSR_21600/DS0008/21600-0008-Data.rda")
wave03 <- da21600.0008
wave03$wave <- 3
rm(da21600.0008)
w3 = subset(wave03, select = c(AID, wave))

#4th wave
load("~/NAS_Share/HP/AddHealthPublic/ICPSR_21600/DS0022/21600-0022-Data.rda")
wave04 <- da21600.0022
wave04$wave <- 4
rm(da21600.0022)
w4 = subset(wave04, select = c(AID, wave))

# 5th wave
load("~/NAS_Share/HP/AddHealthPublic/ICPSR_21600/DS0032/21600-0032-Data.rda")
wave05 <- da21600.0032
wave05$wave <- 5
rm(da21600.0032)
w5 = subset(wave05, select = c(AID, wave))
```

The complete list of respondents can be obtained by aggregating all WX datasets and then getting the unique AID.

-   The numbers of cases for both occasions are the same. It looks like the first wave contains all respondents - no additional respondents were added.
-   To check this observation, we will see both AID from the first wave and All matched.
-   Because the n didn't change, we confirmed that WAVE01 contains all respondents of the study.

```{r}
AH01 <- unique(subset(rbind(w1, w2, w3, w4, w5), select = c(AID)))
test01 <- cbind(wave01, AH01, by="AID")
```

### Generate a complete framework with *AID* and *wave* (Optional)

I personally prefer working with a "complete framework" containing all *AID* and *wave*.

```{r}
lf01 <- rbind(w1, w2, w3, w4, w5)
lf <- lf01[order(lf01$AID, lf01$wave), ]
```

Here is the merged ("stacked") dataset.

```{r}
head(lf01)
```

Sorting by *AID* and *wave*, we can easily identify the data structure by *AID* and *wave*. This **lf** dataset is what I call a "framework" of this data source, which is the one that will be used whenever combining or merging datasets.

```{r}
head(lf)
```

Keep the number of cases (n = `r count(lf)`) for your record. This number should be the number you expect whenever you merge or stack datasets.

```{r}
count(lf)
```

### Time-variant variables from each wave

In this practice, we will select and rename self-rated health (for all 5-wave) and appetite (only for $1^{st}$ and $2^{nd}$ waves) measures.

```{r}
srh1 <- wave01 %>%
  dplyr::select(AID,
         wave,
         "a_srh" = H1GH1,
         "a_poorappetite" = H1FS2)

srh2 <- wave02 %>%
  dplyr::select(AID,
         wave,
         "a_srh" = H2GH1,
         "a_poorappetite" = H2GH22)

srh3 <- wave03 %>%
  dplyr::select(AID,
         wave,
         "a_srh" = H3GH1)

srh4 <- wave04 %>%
  dplyr::select(AID,
         wave,
         "a_srh" = H4GH1)

srh5 <- wave05 %>%
  dplyr::select(AID,
         wave,
         "a_srh" = H5ID1)
```

Please note that how to name the "temporary" datasets. I found that using the combination of 'variable name + wave' minimizes any confusions later.

The 'rbind' function requires all datasets have a same numbers of columns. 'setDT' and 'fill=TRUE' are the functions from a 'data.table' package that override this requirement.

Now, we have created a *long-form dataset (i.e., srh_TV) from five sets of cross-sectional datasets*.

```{r}
srh_TV01 <- rbind(setDT(srh1), setDT(srh2), setDT(srh3), setDT(srh4), setDT(srh5), fill=TRUE)
srh_TV <- srh_TV01[order(srh_TV01$AID, srh_TV01$wave), ]
# 6504, 4834, 4882, 5114, 4196, 25530
```

```{r}
head(srh_TV)
```

### Time-invariant

By definition, when a variable is time-invariant, only one measure from any variable should be applied to all other waves. In this example, we select *sex* from the first wave (because of completeness), which will be applied to the whole long-form dataset.

```{r}
demo_TI <- wave01 %>%
  select(AID,
         "a_sex" = BIO_SEX)
```

### Merging datasets

Once you have selected, created, and modified all required variables by waves, stacking all waves datasets will generate a long-form dataset per wave-person as long as you have keep *AID* and *wave* variables for all datasets.

In this example, we've created three datasets - **lf** (a framework), **demo_TI** (time-invariant), and **srh_TV** (time-variant).

- Framework + time-invariant (i.e., **lf** (a framework) and **demo_TI** (time-invariant))

```{r}
Final01 <- merge(lf, demo_TI, by = c("AID"))
head(Final01)
```

- Framework + time-invariant + time-variant (i.e., **Final01** + **srh_TV** (time-variant))

```{r}
Final02 <- merge(Final01, srh_TV, by = c("AID", "wave"))
head(Final02)
```

### Define *event*, *enter*, and *exit*

The *event* can be defined as your outcomes. Depending on the nature of outcomes, it could be a multiple or repetitive events, requiring more complex survival modeling with more assumptions. 

- Using lag/lead(wave)

```{r}
Final03 <- Final02 %>% 
  group_by(AID) %>% 
  dplyr::mutate(
    enter = lag(wave),
    exit = wave
    ) %>% 
  ungroup()

Final03$enter[Final03$wave == 1 & is.na(Final03$enter)] <- 0
```

Because we used *wave* as an example, it may look more complicated than necessary - for example, we may simply use `enter = exit - 1`. However, this lag/lead function is required when working with the actual date which interval is not always equal to 1. 


