# Survival models: specification, estimation, and interpretation

```{r, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(eha)
library(survival)
library(data.table)
library(flextable)
library(survminer)
library(ggfortify)
library(ggplot2)
library(data.table)
```


```{r, include=FALSE}
oldmort01 <- oldmort
```

Let's think some some feasible models addressing how the survival varies by sex, region, and infant mortality of the cohort, using **oldmort01** dataset. 

Here are some possible models depending on the outcome:

- Descriptive models for survival time

- Linear or Poisson regression on the 'survival time', which can be defined as the time of death (i.e., 'exit'). We may need to subset only those who died, potentially resulting in considerable loss of data.

- Logistic regression for the event, death. How would you incorporate "survival time" in this model?

- Semiparametric survival regression models

- parametric survival regression models

## Nonparametric models

Let's fit Kaplan-Meier (KM) and Nelson-Aalen (NA) estimators using the **oldmort01** dataset from the **eha** package.

- Kaplan-Meier (KM) survival estimator

```{r}
## KM
bysex_KM <- survfit(Surv(enter, exit, event) ~ sex,
                    data=oldmort01,
                    conf.type="log-log")

## Tables
bysex_KM

##summary(bysex)
summary(bysex_KM, times=c(60, 65, 70, 75, 80, 85, 90, 95, 100, 105, 110)) # add time points
```

- Nelson-Aalen (NA) estimator

```{r}
## NA
bysex_NA <- survfit(Surv(enter, exit, event) ~ sex,
                    data=oldmort01,
                    conf.type="log-log",
                    type="fh") # an option for NA estimator

## Tables
bysex_NA

##summary(bysex)
summary(bysex_NA, times=c(60, 65, 70, 75, 80, 85, 90, 95, 100, 105, 110)) # add time points
```

- Overall survival and hazard curves for the population

```{r}

## Plots
par(mfrow = c(1, 2))# Two panels, "one row, two columns".
with(oldmort01, plot(Surv(enter, exit, event), fun = "cumhaz", 
                main = "Cumulativa hazards function",
                xlab = "Duration"))
with(oldmort01, plot(Surv(enter, exit, event),
                main = "Survival function",
                xlab = "Duration"))
```

- Comparison between Male and Female

```{r}
# Plots
par(mfrow = c(1, 2))# Two panels, "one row, two columns".

plot(bysex_KM,
     ylab = "Survival probability",
     xlab = "Time",
     mark.time = T,
     main="Kaplan-Meier survival curve")
legend("topleft", c("Male","Female"),
       lty=c("solid","dashed"), 
       col=c("black","red"))
#abline(h = 0.5, col = "sienna", lty = 3)

plot(bysex_NA,
     ylab = "Cumulative hazard",
     xlab = "Time",
     mark.time = T,
     fun="cumhaz",
     main="Nelson-Aalen cumulative hazard curve")
legend("topleft", c("Male","Female"),
       lty=c("solid","dashed"), 
       col=c("black","red"))
#abline(h = 0.5, col = "sienna", lty = 3)
```
- For a better plot for comparisons

```{r}
library(ggfortify)
library(ggplot2)

autoplot(bysex_KM,
     ylab = "Survival probabilities",
     xlab = "Survival time",
     #mark.time = T,
     main="Kaplan-Meier survival curve estimate with 95% CIs"
     )
```

## Semi-parametric models: Cox Regression


### Model specification
$$h(t)=h_0 (t)\exp(b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR})$$

### Estimation


```{r}
oldmort_cox <- coxreg(Surv(enter, exit, event) ~ sex + region + imr.birth, 
              data = oldmort01) 

print(summary(oldmort_cox), digits = 4)

b_cox <- coef(oldmort_cox)
expb_cox <- exp(coef(oldmort_cox))
```

```{r}

# Plots
par(mfrow = c(1, 2), las = 1)
plot(oldmort_cox, 
     fn = "sur", main = "", 
     #xlab="Duration (year)", 
     ylab="Survival",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
plot(oldmort_cox, 
     fn = "cum", main = "", 
     #xlab="Duration (year)", 
     ylab="Hazard",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
```

To ease interpretation, we exponentiate coefficients (and CIs).

```{r}
exp(coef(oldmort_cox))
```




### Interpretations

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the difference between *coefficients* and $\exp$(*coefficients*)? Specify the metric.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* What is the interpretation when 1) $\exp(b_i) = 1$, 2) $\exp(b_i) < 1$, or 3) $\exp(b_i) > 1$?
* How would you compare $p(death)$ between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated $p(death)$ for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated $p(death)$ for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?


## Logistic regression

To fit logistic regression, 'death' variable was created.

```{r}
oldmort01$death <- ifelse(oldmort01$event == "TRUE", 1, 0)
```


### Model specification

$$ \ln \left( \frac{p(y)}{1-p(y)} \right) = b_0 + b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR}$$



### Estimation

Logistic model was fitted as below.

```{r}
oldmort_log <- glm(death ~ sex + region + imr.birth,
                  data=oldmort01, 
                  family = binomial(link = "logit"))

summary(oldmort_log)
b_log = coef(oldmort_log)
expb = exp(coef(oldmort_log))

```


To ease interpretation, we exponentiate coefficients (and CIs).

```{r}
exp(coef(oldmort_log))
```


### Interpretation

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the difference between *coefficients* and $\exp$(*coefficients*)? Specify the metric.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* What is the interpretation when 1) $exp(b_i) = 1$, 2) $exp(b_i) < 1$, or 3) $exp(b_i) > 1$?
* How would you compare $p(death)$ between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated $p(death)$ for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated $p(death)$ for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?

## Linear regression

### Model specification
$$ Y_{Time\;to\; death} = b_0 + b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR} $$

### Estimation

To fit linear model, we need to subset data for the death and use 'exit' as an outcome.

```{r}
oldmort02 <- oldmort01[oldmort01$death == 1,]
```

```{r}
oldmort_lm <- glm(exit ~ sex + region + imr.birth,
                  data=oldmort02, 
                  family = "gaussian")
summary(oldmort_lm)
b_lm = coef(oldmort_lm)
```


### Interpretation

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* How would you compare the *time* to death between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated time to death for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated time to death for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?


## Weibull model

### Model specification
$$h(t)=h_0 (t)\exp(b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR})$$
The full hazard function for the Weibull PH model is
$$h(t)=\exp(b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)pt^{p-1}$$
Therefore, in terms of $S(t)$,
$$ S(t)=\exp(-(b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)t^p) $$
$p \; (0<p)$ is a shape parameter.

### Estimation


```{r}
# Models
oldmort_wei <- phreg(Surv(enter, exit, event) ~ sex + region + imr.birth, 
              data = oldmort01,
              dist = "weibull")

# Table
#print(summary(oldmort_wei), digits = 4)
oldmort_wei

b_wei <- coef(oldmort_wei)
expb_wei <- exp(coef(oldmort_wei))


# Plots
par(mfrow = c(1, 2), las = 1)
plot(oldmort_wei, 
     fn = "sur", main = "", 
     #xlab="Duration (year)", 
     ylab="Survival",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
plot(oldmort_wei, 
     fn = "cum", main = "", 
     #xlab="Duration (year)", 
     ylab="hazard",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
```

To ease interpretation, we exponentiate coefficients (and CIs).

```{r}
exp(coef(oldmort_wei))
```


### Interpretations

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the difference between *coefficients* and $\exp$(*coefficients*)? Specify the metric.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* What is the interpretation when 1) $\exp(b_i) = 1$, 2) $\exp(b_i) < 1$, or 3) $\exp(b_i) > 1$?
* How would you compare $h(time\;to\;death)$ between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?


## Exponential model

### Model specification
$$h(t)=h_0 (t)\exp(b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR})$$
Exponential model is a specific case of Weibull family when $p$=1.

The full hazard function is
$$h(t)=\exp(b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)pt^{p-1}=\exp(b_0 + b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)$$

Therefore, in terms of $S(t)$,
$$ S(t)=\exp(-(b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)t^p)=\exp(-(b_0 + b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)t) $$

### Estimation


```{r}
# Models
oldmort_exp <- phreg(Surv(enter, exit, event) ~ sex + region + imr.birth, 
                     shape=1, 
                     data = oldmort01,
                     dist = "weibull")

# Table
#print(summary(oldmort_wei), digits = 4)
oldmort_exp

b_exp <- coef(oldmort_exp)
expb_exp <- exp(coef(oldmort_exp))


# Plots
par(mfrow = c(1, 2), las = 1)
plot(oldmort_exp, 
     fn = "sur", main = "", 
     #xlab="Duration (year)", 
     ylab="Survival",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
plot(oldmort_exp, 
     fn = "cum", main = "", 
     #xlab="Duration (year)", 
     ylab="hazard",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
```

To ease interpretation, we exponentiate coefficients (and CIs).

```{r}
exp(coef(oldmort_exp))
```


### Interpretations

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the difference between *coefficients* and $\exp$(*coefficients*)? Specify the metric.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* What is the interpretation when 1) $\exp(b_i) = 1$, 2) $\exp(b_i) < 1$, or 3) $\exp(b_i) > 1$?
* How would you compare $h(time\;to\;death)$ between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?


## Gompertz model

### Model specification
$$h(t)=h_0 (t)\exp(b_1\times D_f + b_2 \times D_{ind} + b_3 \times D_{rural} + b_4 \times X_{IMR})$$

>Gompertz model is characterized by an exponentially increasing hazard function with fixed rate $r$ ($-\infty < r < \infty$).
>when $r < 0$, the hazard function $h$ is decreasing “too fast” to define a proper survival function, and $r=0$ gives the exponential distribution as a special case. And for each fixed $r$, the family of distributions indexed by $p > 0$ constitutes a proportional hazards family of distributions, and the corresponding regression model is written as
>Göran Broström, https://cran.r-project.org/web/packages/eha/vignettes/gompertz.html

$$h(t)=\exp(b_1 x_1 + b_2 x_2 + \cdots + b_n x_n)pe^{rt}$$


### Estimation


```{r}
# Models
oldmort_gomp <- phreg(Surv(enter, exit, event) ~ sex + region + imr.birth, 
              data = oldmort01, 
              dist = "gompertz")

# Table
#print(summary(parm), digits = 4)
oldmort_gomp

b_gomp <- coef(oldmort_gomp)
expb_gomp <- exp(coef(oldmort_gomp))

# Plots
par(mfrow = c(1, 2), las = 1)
plot(oldmort_gomp, 
     fn = "sur", main = "", 
     #xlab="Duration (year)", 
     ylab="Survival",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
plot(oldmort_gomp, 
     fn = "cum", main = "", 
     #xlab="Duration (year)", 
     ylab="hazard",
     #xlim=c(0, 1) 
     #ylim=c(ymin, ymax)
     )
```
To ease interpretation, we exponentiate coefficients (and CIs).

```{r}
exp(coef(oldmort_gomp))
```


### Interpretations

* What is the metric of $y$ and $b_i$, respectively? 
* Interpret $b_0, b_1,$ and $b_2$, respectively.
* What is the difference between *coefficients* and $\exp$(*coefficients*)? Specify the metric.
* What is the interpretation when 1) $b_i = 0$, 2) $b_i < 0$, or 3) $b_i > 0$?
* What is the interpretation when 1) $\exp(b_i) = 1$, 2) $\exp(b_i) < 1$, or 3) $\exp(b_i) > 1$?
* How would you compare $h(time\;to\;death)$ between two groups of people below? Is the effect additive or multiplicative?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 0, and IMR = 0 vs. those who with sex = 1, region = 0, and IMR = 0?
  - What is the estimated $h(time\;to\;death)$ for those who with sex = 0, region = 2, and IMR = 90 vs. those who with sex = 1, region = 2, and IMR = 90?



## Graphs

The following figures summarize cumulative hazard curves by different survival models.

```{r}
# Plots
par(mfrow = c(2, 2), las = 1)

plot(oldmort_cox, 
     fn = "cum", 
     main = "Cox", 
     #xlab="Duration (year)", 
     ylab="",
     #xlim=c(0, 1) 
     ylim=c(0, 10)
     )

plot(oldmort_wei, 
     fn = "cum", 
     main = "Weibull", 
     #xlab="Duration (year)", 
     ylab="",
     #xlim=c(0, 1) 
     ylim=c(0, 10)
     )

plot(oldmort_exp, 
     fn = "cum", 
     main = "Exponential", 
     #xlab="Duration (year)", 
     ylab="",
     #xlim=c(0, 1) 
     ylim=c(0, 10)
     )

plot(oldmort_gomp, 
     fn = "cum", 
     main = "Gompertz", 
     #xlab="Duration (year)", 
     ylab="",
     #xlim=c(0, 1) 
     ylim=c(0, 10)
     )
```

